Notes from the President

### 2013-10-16 pnsac/newsletter/article_Feature1.tex
Changed directory name. Primary newsletter files now in newsletter/
October 2013 newsletter (DRAFT)

\title{My Days at the Home of the Merlin Engine}
\author{Richard Lodge}

@@

As a new employee, part of my Rolls-Royce induction process was to get
to know some of the production and other R-R facilities around
Derby. After a few weeks of working in the Overheads Department, it
was suggested that it was time for me to go visiting. I was happy to
do this because any opportunity to leave the main accounting office
for a time was to be jumped at.

Rolls-Royce had an internal bus service to take employees between
sites. To catch the bus, an employee went down to the door of the
building and waited for a bus to turn up -- no booking or written
schedule.  I remember my first trip on a company bus. It took me about
half an hour of asking people around how to get to the site I
wanted. I was beginning to realize that much of the operations of R-R
were not written and were passed around by word of mouth and the long
time employees passing on information to the newcomers. Years later,
when working in Northern Canada in native communities I once again
came across this way of handing down unwritten information from the
Elders to the younger people. Before starting my job at R-R I had
believed that the company would be highly organized and disciplined.

I finally managed to find my way to the building I was due to visit.
It was the Foundry. I was introduced to the men who worked in this
evil smelling place. Each of the major engine parts was individually
cast -- there was no automation and a good foundry man could put
together the sand and other parts needed for the casting, mostly from
memory. During WW2 the foundry had concentrated entirely on the
production of military engines, mostly Merlins.  By the time I was
shown round the place there was a bewildering conglomeration of parts
being made for piston, turbo prop and jet engines.

As a young accountant I was expected to understand how the paperwork
and accounting for all this worked. It was a daunting thought and I
left the Foundry feeling overwhelmed.  Throughout my time at R-R I
never became comfortable dealing with the Foundry -- everything seemed
to be so imprecise and up to the skill on the foundry man. Each time I
went there I tried to find out how the men knew how to produce the
same casting every time and the answer was always the same: "Oh, we
just know what to do".  Many of these comments were probably made to
try to make a young guy in a suit from the office look foolish but
they were also largely true!

Sitting in the Overheads Department I saw many of the invoices for
purchases. I could hardly believe it when I saw invoices going through
for payment for dozens of chickens. The delivery was never to the
works canteen but always to the R-R site at Hucknall, a short distance
outside Derby. My supervisor told me that the chickens were fired at
engines on test. I did not believe him until I managed to find a
pretext to visit Hucknall. After another challenge finding the works
bus to get me there, I arrived at the site and was shown round and all
was revealed. Dead chickens ending their time on earth being used to
test the strength of compressor blades in an engine on full power
seemed to me to be a very imprecise way of testing a jet
engine. Another lesson learned -- not everybody was trying to make me
look foolish all the time.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{bird_gun.eps}
   %caption of the figure 
   \caption*{\small \em A 1960s chicken gun as used at Hucknall .}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:birdgun}
\end{figure}

The visits to the various R-R sites around Derby took several months
to complete. I had to have a good reason to leave my office and go
visiting, particularly as I was now working in a departmental
position. Eventually another bus trip to the large new complex at
Sinfin was required or I somehow managed to talk my way into finding
it necessary to go there. Sinfin, among many other things, housed the
main engine test beds and I was curious to find out what happened
there. Unlike the evil smelling Foundry, Sinfin was clean, new and had
plenty of windows.

The men in the test shop told me about the engine testing process. I
was still new to R-R and had a lot to learn. After a while I was asked
if I would like to see an engine on test. Of course I said yes; it
seemed interesting. With little experience I did not realize that an
aero engine on test is nothing much to see. One of the men said he
would take me to a Conway engine on test. He said he would open the
test bed door to let me see the engine and that I would find it
somewhat noisy. He stood behind the door, opened it and let me look
inside at the engine on full power. I now wear hearing aids. Health
and safety were two words not used in those days when many of the men
were used to serving in the war. This was 1960s Britain.

I have always enjoyed travelling and would jump at any opportunity to
travel to a new place. After a few months it was suggested to me that
I should visit the Scottish factory at Hillington near Glasgow. I
asked how I should get there. ``Oh. You might as well fly. There is a
flight every day that leaves in the morning and comes back in the
afternoon.'' I had never flown before.

The UK being a small country, everybody went around by train.  As
usual, I was given very little information and told to get myself to
East Midlands airport early on the appointed morning, where a ticket
would be waiting for me. On this occasion I had the good sense to take
my own car to the airport and did not get lost. I expected to find a
plane waiting, painted in some sort of Rolls-Royce colours. Instead,
after some questioning, I found out that I would be travelling on a
small airline, called British Midland, which ran a service, mainly for
R-R personnel, between the Derby East Midlands Airport, near Derby and
Glasgow factories Airport. Far from being in some interesting plane
connected with R-R, I found that our plane was an old DC3 which had
seen better days.I duly got on the plane and was somewhat nervous
flying on my first flight.

An hour later we arrived in Scotland, also my first trip over the
Border, and I spent the day being shown around the factories and
offices.  I remember very little about the day but vividly remember
the flight back to Derby. It must have been in the winter because it
was getting dark as I boarded the plane around 5.00 pm. Seating was
first come first served and all the old sweats travelling knew what to
do. I ended up on the back row near the cargo area. The main thing I
remember was the crew having a fierce argument about the load on the
plane which was full. The discussion centred around the amount of
cargo which could be taken on the flight. I was quite relieved to get
back to Derby in one piece.

In the next article I will describe how I worked hard to climb the
hierarchy at R-R and learnt in an uncomfortable way some lessons on
people management.

>>>>

\title{My Days at the Home of the Merlin Engine}
\subtitle{Part 3}
\author{Richard Lodge}

@@

``Lodge, the expenses on the charity wagon go to Miss Prod''. What was a
charity wagon and who was Miss Prod? I was working in the Overheads
Department of the Rolls-Royce Aero Engine Division in Derby. As a
young, ambitious accountant, I did not want to admit to the people in
the department that I did not know what they were talking about. I was
conscious that I had spent several weeks of initiation under the
guidance of the old timer, Les Hart, and should now know something
about the company. I was sitting in front of a huge leather bound
ledger where accounting information was recorded. Eventually, by
listening to other people talking around me, I established that a
charity wagon was a much prized company car and that Miss Prod was not
a person but referred to ``miscellaneous production'' work at R-R. Put
in plain English, the sentence meant that the cost of repairing this
individual's car was to be charged to a special account in the
miscellaneous production section of the overheads. There was so much
to learn!  I was completely bewildered. Each day was a challenge and
the thought that aero engines were produced in the surrounding
factories seemed very far away.

On being placed in the Overheads Department, I was given an internal
mail address. The company was so large that a worldwide internal mail
system had been established and I rapidly found out that I was at the
bottom of the heap. My address was JD/JS/LOD which meant that JS was
my immediate supervisor and JD was his boss. If you were important in
the company, your mail would come to you with just your initials such
as JD. Everybody with any ambition wanted to get a single initials
mailing address. If you were really close to God, such as a director
of the company, your mailing address would be Psn for the CEO, Sir
Denning Pearson, or DPH for David Huddie, the Director of Engineering.

There were continuous battles between the engine development engineers
and the accountants. When the engineers spent money, they tried
wherever possible to avoid having it charged against their
budget. Their favourite game was to charge the cost into the overheads
where it would be lost in a mass of other expenditure such as emery
paper and toilet rolls. Our task was to catch them at their game and
reallocate the expense back to them. The development engineers were
very skilled at this part of their work and became so good at it that
they were able to hide major cost overruns, sometimes
indefinitely. Little did I realize that I was helping to perpetuate
the culture that was one of the causes of R-R going bust a few years
later.

In the 1960s R-R was a paternalistic organization, steeped in
tradition, still operating with many cost-plus government contracts,
vast amounts of paper and virtually no computers. The Ministry of
Aviation would regularly send auditors into the company to check how
public money was being spent. These gentlemen had a permanent office
allocated to them and would regularly attack the expenses recorded by
the Overheads Department. When this happened, we and the development
engineers became great buddies for a short time. Our unspoken job was
to confuse the government auditors as much as possible so that they
would not report back to their masters that things were not as they
should be. We became quite good at our confusion duties. It was not a
difficult job because we were frequently confused by all the paper and
numbers as well.

R-R continued to deliver new aero engines to customers and I gradually
started to understand what I was doing. After a few months, my
supervisor was promoted and I was appointed supervisor of the
Overheads Department, much to the disgust of many of the people in the
department. Although my pleasure was to be short-lived, I was
delighted to get this promotion. By the time of my promotion, I had
been around many of the facilities of R-R both in the Derby area and
in Scotland. I felt that I knew my way around the company.

Almost everyone working in the Overheads Department, including the
three section leaders under me, were older and more experienced than
I. They did not like this young upstart accountant becoming their
boss. Underground warfare started between us. Several of the men had
fought through WW2 and were well practiced at the gentle art of ``dumb
insolence''. The three section leaders came to me one day and told me
they really disliked me being their boss. This was not good for my
ego. I suggested to them that the four of us should go out after work
and have a beer and that on this occasion I would just be one of them
and not their boss. I bought the first round of beer and they really
tore into me. After several more rounds of beer, we agreed to a truce
and decided on a new strategy. I would stop trying to micromanage them
and they would stop their dumb insolence. Of course, they never
admitted that they were practicing dumb insolence. Life became much
better after our beers. Miss Prods, charity wagons, drop-offs and all
the other R-R jargon became normal for me with the active co-operation
of my three section leaders.

Apart from the pay rise upon my promotion I had now moved up in the
company mailing system and my address had become JD/LOD. This probably
helped my head to swell until the three section leaders brought me
rapidly back to earth.

I must have learnt a few of life's lessons about people management
because when I was again promoted, my three section leaders tried to
encourage me to refuse the promotion. No chance. I was once again
moving up the mailing system and would in future have my mail
addressed to LOD, without a more senior person's initials appearing
before mine. I was going in the right direction.

### 2013-10-16 pnsac/newsletter/article_PresReport.tex
October 2013 newsletter (DRAFT)

Canada Day is now behind us.  We had an excellent day with a good
number of our volunteers showing off the North Star to the general
public. Judging by the activity around the plane all the day, we have
now developed a display format which is interesting to visitors.
Canada Day was the only day this year when the plane was on display to
the general public.

On June 27, in co-operation with the Museum, who hosted the event, we
showed off the plane to a group of donors attending the Donor
Appreciation Evening.  This was a successful evening which is likely
to be repeated.

In my last Notes from the President, I referred to plans for
celebrating the 10th anniversary of the founding of the Project North
Star Association.  This should have taken place around June 24.
Unfortunately, the event passed without us being able mark the
day. This was mainly due to situations beyond our control.

Our 2013 AGM was held on June 8, and all the existing directors were
re-elected.  This will enable the directors, who have now become
experienced, to concentrate on long-term matters.

The actual work on the restoration of the plane is going very well, as
reported elsewhere in this newsletter. The major issue facing the
Board in the next year will be to consider how to secure the future of
PNSAC, both in terms of volunteers and the funding of restoration
work.

We have now reached a point where it is necessary to update our
Memorandum of Understanding (MOU) with the Museum and our work plan
for completing the restoration of the plane.  Our volunteers have
already worked over 60,000 hours on the project and we expect to work
a similar number of hours before the restoration is completed.

There have been several informal meetings to discuss our future
operations and the opportunities available for new volunteers.  Within
the next few weeks we are expecting to meet with the Director General
of the Aviation Museum and members of his staff to work on the MOU
update.

Later this month we are meeting with senior management of the Museum
Foundation, the body charged with corporate fundraising for the Canada
Science and Technology Museums Corporation, to discuss how PNSAC can
work with the Foundation to raise funds for the North Star restoration
and the museums in general.

With the initiatives above, we are hoping in the next few months to
consolidate our position as an important organization working with the
Aviation Museum

>>>>

During the past three months we have moved steadily forward with the
restoration of the North Star.  Our progress has been somewhat held up
by lack of available volunteers.

Our Association has two distinct and very important sides.  First and
foremost we are a group whose interest is in the North Star and the
restoration and preservation of the plane at the Museum.  Secondly,
over the years, we have developed a strong social side to the
Association where volunteers and other members can join in interesting
events centered around aviation. Elsewhere in this issue of the
Chronicle you will find a description by Bill Tate our Vice President,
of the very successful trip on October 4 to the Navcan Air Control
Centre and to the Bombardier factory in Montr\'{e}al.

After 10 years of restoration work on the North Star we recognize that
we must review our progress and develop a plan for the rest of the
work on the plane.  There will be three aspects to this review.  Our
first task will be to identify the major sections of the project and
develop ways of actually carrying out the restoration.  The
undercarriage of the plane is a particularly complicated challenge.
We then need to prepare a plan with projected timelines for completing
each part of the work.  Our third task will be to estimate costs and
to develop ideas for financing it.  During this process we must also
ensure that we will have sufficient qualified volunteers to get it
done.

Before the end of November we are hoping to have meetings with the
senior management of the Museum to update planning of all the tasks
above.  The end result of this should be a documented project plan and
a new Memorandum of Understanding between the Canada Aviation and
Space Museum and the Project North Star Association of Canada.

At times our progress on the restoration of the North Star may appear
to Association members to be somewhat slow.  Unlike other
organizations in Canada such as Vintage Wings in Gatineau or the
Canada Warplane Heritage Museum in Hamilton, PNSAC cannot make
decisions about the future of the North Star on its own.  As an
Association working with a Federal Government institution we are often
affected by decisions not related to the work we do.  This can at
times provide interesting challenges.

In conclusion, I would say that the Directors are working hard with
the Museum management to move our project forward even though at times
it may appear to be very difficult.  I hope that I will be able to
report significant progress in the next issue of the Chronicle.  In
the meantime we have some interesting events in the planning stages on
the social side of the Association.  Stay tuned.
